---
title: "Nonsynonymous substitutions loser>gainer as a function of VAF: excess of deleterious variants among recent variants?"
author: "Konstantin Popadin"
date: "10 Oct 2022"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls(all=TRUE))
library(knitr)
library(tidyverse)
library(kableExtra)
library(vioplot) # install.packages('vioplot')
```

## Background
<br>
We suspect that fraction of deleterious variants (loser>gainer) will be higher among recent (low VAF) variants
<br>
<br><br>

## 1. Read file and touch the data
***
```{r}
AllNons = read.table("../new_data/data/cov_glue_analys.csv", sep = ',' ,header = TRUE)
dim(AllNons)
names(AllNons)
head(AllNons)
table(AllNons$MutType)
table(AllNons$mut_status_char)
table(AllNons$mut_status)
AllNons$mut_status[AllNons$mut_status == -1] <-0 
table(AllNons$mut_status) # == 0 if it is C>U or G>U and == 1 if something other
summary(AllNons$Proportion)
str(AllNons$Proportion)
str(AllNons$mut_status)
```
<br><br><br>

## 2. run very basic statistics: if number of loser>gainer >> gainer>loser
***
```{r}
nrow(AllNons[AllNons$mut_status_char == 'C>U G>U',]) # 2897
nrow(AllNons[AllNons$mut_status_char == 'U>C U>G',]) # 390
nrow(AllNons[AllNons$mut_status_char == 'unknown',]) # 2065
```
<br><br><br>

## 3. run basic statistics: if frequency of loser>gainer (==1) from ALL nonsynonymous mutations is associated to VAF?
***
```{r}
summary(glm(AllNons$mut_status ~ AllNons$Proportion, family = 'binomial'))
cor.test(AllNons$mut_status,AllNons$Proportion, method = 'spearman')
boxplot(AllNons[AllNons$mut_status == 1,]$Proportion,AllNons[AllNons$mut_status == 0,]$Proportion, outline = FALSE)
wilcox.test(AllNons[AllNons$mut_status == 1,]$Proportion,AllNons[AllNons$mut_status == 0,]$Proportion)
t.test(AllNons[AllNons$mut_status == 1,]$Proportion,AllNons[AllNons$mut_status == 0,]$Proportion)

AllNons = AllNons[order(-AllNons$Proportion),] # start from high proportions: old mutations and move to recent ones
for (i in 0:(as.integer((nrow(AllNons)/30))-1)) # as.integer(5.6)
{
temp =  AllNons[seq((i)*30,(i+1)*30-1,1),]
fr = nrow(temp[temp$mut_status == 1,])/nrow(temp)
Out = data.frame(i,fr)
if (i == 0) {FinalOut = Out}
if (i >  0) {FinalOut = rbind(FinalOut,Out)}
}
cor.test(FinalOut$i,FinalOut$fr, method = 'spearman')
plot(FinalOut$i,FinalOut$fr) # after i == 25 (row = 25*30 = 750) there is a robust decrease in the loser>gainer mutations. Why? Can we check it by logistic regression?

summary(glm(AllNons[c(750:nrow(AllNons)),]$mut_status ~ AllNons[c(750:nrow(AllNons)),]$Proportion, family = 'binomial')) # very strong positive
summary(glm(AllNons[c(0:750),]$mut_status ~ AllNons[c(0:750),]$Proportion, family = 'binomial')) # strong negative

```  
<br><br><br>

## 4. run specific statistics: if frequency of loser>gainer (==1) as compared to gainer>losers is associated to VAF?
### despite the VACT that trend is the same (decrease in loser>gainer with time) the probability is still > 50%
***
```{r}
AllNons = AllNons[AllNons$mut_status_char != 'unknown',] # loser>gainer versus gainer>loser
# AllNons = AllNons[AllNons$mut_status_char == 'unknown' | AllNons$mut_status_char == 'C>U G>U',] # loser>gainer versus others 'unknown'
summary(glm(AllNons$mut_status ~ AllNons$Proportion, family = 'binomial'))
cor.test(AllNons$mut_status,AllNons$Proportion, method = 'spearman')
boxplot(AllNons[AllNons$mut_status == 1,]$Proportion,AllNons[AllNons$mut_status == 0,]$Proportion, outline = FALSE)
wilcox.test(AllNons[AllNons$mut_status == 1,]$Proportion,AllNons[AllNons$mut_status == 0,]$Proportion)
t.test(AllNons[AllNons$mut_status == 1,]$Proportion,AllNons[AllNons$mut_status == 0,]$Proportion)

AllNons = AllNons[order(-AllNons$Proportion),] # start from high proportions: old mutations and move to recent ones
for (i in 0:(as.integer((nrow(AllNons)/30))-1)) # as.integer(5.6)
{
temp =  AllNons[seq((i)*30,(i+1)*30-1,1),]
fr = nrow(temp[temp$mut_status == 1,])/nrow(temp)
Out = data.frame(i,fr)
if (i == 0) {FinalOut = Out}
if (i >  0) {FinalOut = rbind(FinalOut,Out)}
}
cor.test(FinalOut$i,FinalOut$fr, method = 'spearman')
plot(FinalOut$i,FinalOut$fr) # after i == 20 (row = 20*30 = 600) there is a robust decrease in the loser>gainer mutations. Why? Can we check it by logistic regression?

summary(glm(AllNons[c(600:nrow(AllNons)),]$mut_status ~ AllNons[c(600:nrow(AllNons)),]$Proportion, family = 'binomial')) # strong positive
summary(glm(AllNons[c(0:600),]$mut_status ~ AllNons[c(0:600),]$Proportion, family = 'binomial')) # negative

```

